Longitudinal Principal Manifold Estimation

### 2023-10-06 paper/main.tex

Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively. For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g. the surface of the 3D brain region, at time $t$. 
\begin{definition}
  \label{def:lpme} Given a random $D$-vector $\boldsymbol{X}_t$ observed over time, $\boldsymbol{X}$ denotes the collection of data clouds observed at all time points, such that the manifold $M_F^{d+1}$ is representative of $\boldsymbol{X}$ and for given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows. 
\begin{align}
  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \label{eq:11} \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \nonumber
.\label{eq:newKappa}
\end{align}
where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\boldsymbol{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}
By minimizing the functional above, we retain the goals of Eq.~\eqref{eq:5} while also penalizing the changes in $F$ over time. Thus, we seek to minimize the distance between the observations $\mathbf{X_t}$ and their projections to $f_t$ while separately discouraging roughness both over the manifolds at each time point and over the changes in manifolds in the time dimension. The ability to distinguish between smoothing at a fixed time point and smoothing over time differentiates this approach from a thin plate spline, which would smooth over both the manifold and time jointly. Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa.

\subsection{Algorithm}

To solve this problem, we propose an extension of the PME algorithm developed in \cite{mengPrincipalManifoldEstimation2021}. The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.

>>>>

Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively. For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g. the surface of the 3D brain region, at time $t$. 
\begin{definition}
  \label{def:lpme} Given a random $D$-vector $\boldsymbol{X}_t$ observed over time, $\boldsymbol{X}$ denotes the collection of data clouds observed at all time points, such that the manifold $M_F^{d+1}$ is representative of $\boldsymbol{X}$ and for given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows. 
\begin{align}\label{eq:newKappa}
  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt. \nonumber
\end{align}
where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\boldsymbol{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}
By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, while imposing smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$. These goals correspond to minizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively. Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions. Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together. Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of roughness shows minimal changes over time, or vice versa. Generally, our proposed framework can be used in applications where a subset of variables of interest is high-dimensional set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered to be smooth over the other set of variables.

\subsection{The LPME Algorithm}

The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.
